---
title: "Mass Spectrometry Protein List Analysis"
description: |
  This R script analyzes and compares protein lists from three mass spectrometry experiments: THY-Tau22 mouse brains, human brains, and WJE3 astrocytes. It maps genes between mouse and human orthologs, handles unmapped genes through manual curation, and generates Venn diagrams to visualize the overlap between protein sets across different experimental conditions.
author:
  - first_name: "Ayush"
    last_name: "Noori"
    url: https://www.github.com/ayushnoori
    affiliation: Massachusetts General Hospital
    affiliation_url: https://www.serranopozolab.org
    orcid_id: 0000-0003-1420-1236
output:
  distill::distill_article:
    toc: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Dependencies

Load requisite packages and define directories. Note that this script may also use my personal utilities package `brainstorm`, which can be downloaded via `devtools::install_github("ayushnoori/brainstorm")`.

```{r load-packages, message=FALSE, warning=FALSE}
# Data manipulation
library(data.table)
library(purrr)
library(magrittr)

# Plotting
library(ggplot2)

# ID mapping
library(biomaRt)

# Relative file paths
library(here)

# Excel output
library(openxlsx)
library(brainstorm)

# Also requires BiocManager::install(c("qusage"))
```

Note that directories are relative to the R project path.

```{r define-directores}
# Set directories
human_WJE3astro_dir = here("Data", "Mass_spec_human_WJE3astro")
mouse_brains_dir = here("Data", "Mass_spec_mouse_brains")
data_dir = here("Data", "Mass_spec_proteins")
results_dir = here("Results", "Mass_spec_proteins")
```

Read gene lists and map from mouse to human and human to mouse.

```{r read-data}
# Read lists
gene_lists = list(
  AD_CTRL = readxl::read_excel(here(human_WJE3astro_dir, "Gfap_IP_protein_lists_for_Ayush_07302024.xlsx"), sheet = "AD+CTRL 1.5-fold GFAP IP vs IgG") %>% as.data.table(),
  MsWt_MtGfapMyc = readxl::read_excel(here(human_WJE3astro_dir, "Gfap_IP_protein_lists_for_Ayush_07302024.xlsx"), sheet = "Ms Wt+Mt Gfap-Myc IP vs. GFP") %>% as.data.table()
)

# Get intersection of gene lists
mouse_gene_list = readxl::read_excel(here(mouse_brains_dir, "List of 89 proteins false for GFP.xlsx"), sheet = "Functional Annotations") %>% as.data.table() %>% .$Gene
human_gene_list = gene_lists$AD_CTRL$Protein
astro_gene_list = gene_lists$MsWt_MtGfapMyc$Protein
astro_gene_list = astro_gene_list[!astro_gene_list %in% c("Igkv8-26", "NA")]

# Convert mouse to human and human to mouse
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")

# mouse to human
mouse_to_human = function(x) {
  return(getLDS(attributes = c("mgi_symbol"), filters = "mgi_symbol", values = x, mart = mouse, attributesL = c("hgnc_symbol"), martL = human, uniqueRows=T))
}

human_to_mouse = function(x) {
  return(getLDS(attributes = c("hgnc_symbol"), filters = "hgnc_symbol", values = x, mart = human, attributesL = c("mgi_symbol"), martL = mouse, uniqueRows=T))
}

mm_genes = list(
  mouse_brains = mouse_gene_list,
  human = unique(human_to_mouse(human_gene_list)[["MGI.symbol"]]),
  astro = astro_gene_list
)

hs_genes = list(
  mouse_brains = unique(mouse_to_human(mouse_gene_list)[["HGNC.symbol"]]),
  human = human_gene_list,
  astro = unique(mouse_to_human(astro_gene_list)[["HGNC.symbol"]])
)

# Print results
# Note that some genes have a 1:many mapping, so the final count may be inflated
# This is corrected by the manual conversion of any unmapped genes
message("Mouse brains: ", length(mm_genes$mouse_brains), " genes --> ", length(hs_genes$mouse_brains), " genes")
message("Human: ", length(hs_genes$human), " genes --> ", length(mm_genes$human), " genes")
message("WJE3 astrocytes: ", length(mm_genes$astro), " genes --> ", length(hs_genes$astro), " genes")
# Mouse brains: 87 genes --> 77 genes
# Human: 100 genes --> 98 genes
# WJE3 astrocytes: 32 genes --> 28 genes

# Get full mapping tables
gene_mappings = list(
  mouse_brains = mouse_to_human(mouse_gene_list),
  human = human_to_mouse(human_gene_list),
  astro = mouse_to_human(astro_gene_list)
)

# Add unmapped genes via manual conversion
# https://www.ncbi.nlm.nih.gov/gene/?term=(Hspd1)+AND+%22Mus+musculus%22%5Bporgn%3A__txid10090%5D
# Search term: (Rpl11) AND "Mus musculus"[porgn:__txid10090] 
# Check orthologs manually in "Summary" or click "Orthologs" > "human"
mouse_brains_not_mapped = mouse_gene_list[!mouse_gene_list %in% gene_mappings$mouse_brains$MGI.symbol]
unmapped_mouse_brains_df = data.frame(
  MGI.symbol = mouse_brains_not_mapped,
  HGNC.symbol = c("FLT3LG", "RAB7A", "RPL11", "RPS3A", "VDAC1", "HSPD1", "ATP5F1B", "TARS1", "RPL17", "H1-4", "HMGA1", "RPS18")
)

# For Gm6133, checked in UniProt: https://www.uniprot.org/uniprotkb/B2RY53/entry
# Searched alias, "Large ribosomal subunit protein uL22" in NCBI Ensembl
# First human hit was RPL17

# For Rps18-ps6, checked in UniProt: https://www.uniprot.org/uniprotkb/F6YVP7/entry
# Searched alias, "Small ribosomal subunit protein uS13" in NCBI Ensembl
# First human hit was RPS18

# Search term: (NEFM) AND "Homo sapiens"[porgn:__txid9606] 
human_not_mapped = human_gene_list[!human_gene_list %in% gene_mappings$human$HGNC.symbol]
unmapped_human_df = data.frame(
  HGNC.symbol = human_not_mapped,
  MGI.symbol = c("Nefm", "Eif4a2", "Rpl9", "Vdac1")
)

# Search term: (Dnaja1) AND "Mus musculus"[porgn:__txid10090]
astro_not_mapped = astro_gene_list[!astro_gene_list %in% gene_mappings$astro$MGI.symbol]
unmapped_astro_df = data.frame(
  MGI.symbol = astro_not_mapped,
  HGNC.symbol = c("DNAJA1", "VDAC1", "MYL12A", "ATP5F1B")
)

# Add manually converted unmapped genes back to mapping tables
gene_mappings$mouse_brains = rbind(unmapped_mouse_brains_df, gene_mappings$mouse_brains)
gene_mappings$human = rbind(unmapped_human_df, gene_mappings$human)
gene_mappings$astro = rbind(unmapped_astro_df, gene_mappings$astro)

# Add experiment column
gene_mappings$mouse_brains[["experiment"]] = "THY-Tau22"
gene_mappings$human[["experiment"]] = "Human brain"
gene_mappings$astro[["experiment"]] = "WJE3 astrocytes"

# Combine final result
IP_proteins = rbind(gene_mappings$mouse_brains, gene_mappings$human, gene_mappings$astro) %>%
  as.data.table() %>%
  setnames(c("MGI.symbol", "HGNC.symbol"), c("mouse", "human"))
```

Compute Venn diagram.

```{r compute-intersection}
# Remove duplicates from each species column for each experiment
mouse_genes = list(
  THY_Tau22 = unique(IP_proteins[experiment == "THY-Tau22", mouse]),
  Human_brains = unique(IP_proteins[experiment == "Human brain", mouse]),
  WJE3_astrocytes = unique(IP_proteins[experiment == "WJE3 astrocytes", mouse])
)

human_genes = list(
  THY_Tau22 = unique(IP_proteins[experiment == "THY-Tau22", human]),
  Human_brains = unique(IP_proteins[experiment == "Human brain", human]),
  WJE3_astrocytes = unique(IP_proteins[experiment == "WJE3 astrocytes", human])
)

# Compute pairwise intersection
mouse_venn = list(
  THY_Tau22_and_Human_brain = intersect(mouse_genes$THY_Tau22, mouse_genes$Human_brain),
  THY_Tau22_and_WJE3_astrocytes = intersect(mouse_genes$THY_Tau22, mouse_genes$WJE3_astrocytes),
  Human_brain_and_WJE3_astrocytes = intersect(mouse_genes$Human_brain, mouse_genes$WJE3_astrocytes),
  All = Reduce(intersect, mouse_genes)
)

human_venn = list(
  THY_Tau22_and_Human_brains = intersect(human_genes$THY_Tau22, human_genes$Human_brain),
  THY_Tau22_and_WJE3_astrocytes = intersect(human_genes$THY_Tau22, human_genes$WJE3_astrocytes),
  Human_brain_and_WJE3_astrocytes = intersect(human_genes$Human_brain, human_genes$WJE3_astrocytes),
  All = Reduce(intersect, human_genes)
)

print_venn = function(pairwise_list) {
  for (name in names(pairwise_list)) {
    clean_name = gsub("_", " ", name)
    cat(clean_name, ": ", paste(pairwise_list[[name]], collapse = ", "), "\n", sep = "")
  }
}

print_venn(human_venn)
```

Create Venn diagram.

```{r venn-diagram}
# devtools::install_github("yanlinlin82/ggvenn")
library(ggvenn)

p = ggvenn(human_genes, show_stats = "c")
```